---
title: "Analyze crime data"
author: "Carsten Behring"
date: '`r Sys.Date()`'
output:
  html_document:
     fig_width: 15
     fig_height: 10
     toc: true
     code_folding: hide

---




```{r}
library(tidyverse)
library(janitor)
library(ggplot2)
library(lubridate)
library(viridis)
library(forcats)

```

# Introduction

<style type="text/css">

 .main-container { width: 90%; max-width:100%;}

</style>

The goal of this report is to show, which times of the day are more dangerous for which type of crimes.
We have data for Seattle and San Francisco and want to analyze the occurrence patterns of type of crimes per time of the day.

# Data source

The data to be used is the criminal statistic of the city of Seattle downloadable here:
https://github.com/uwescience/datasci_course_materials/tree/master/assignment6

## Analysis for Seattle

The data for Seattle looks likes this

```{r}
seattle <- readr::read_csv("./seattle_incidents_summer_2014.csv") %>%
    clean_names()

glimpse(seattle,witdth=120)


```



```{r}
seattle <- 
    seattle %>% mutate(occurred_date_or_date_range_start=mdy_hms(occurred_date_or_date_range_start),
                 start_hour = hour(occurred_date_or_date_range_start))

```

First I plot all crime incidents with their starting hour and the district they occur:

```{r}
ggplot(seattle) +
    geom_jitter(aes(start_hour,summarized_offense_description,color=district_sector),alpha=0.5) +
    labs(x="Hour of day", y="Type of crime",color="District")
``` 

This shows clearly, that the crime of 'Prostitution' occur mainly in one district('N') concentrated at certain time of the day, namely between mid-day and 2 o clock in the morning.

Further, it seems that district R and S have overall higher crime rates.



Different type of crime occur with different frequency:

```{r}
tabyl(seattle,summarized_offense_description,sort = T)

``` 



This plot shows, that car prowls and other different types of theft / robbery occur most often.

```{r}

seattle %>%
    count(summarized_offense_description) %>%
    ggplot() +
    geom_col(aes(x=fct_reorder(summarized_offense_description,n) ,y=n)) +
    coord_flip() +
    labs(y="# crimes",
         x="type of crime",
         title="Frequencies of crime types") 


``` 

To get a nice overview of the type of crime occurring per day, we will look at a heat map which 
shows how many crimes(and of which type) occurred at which time of the day.


```{r}

seattle %>% count(summarized_offense_description,start_hour) %>%
    mutate(total=sum(n)) %>%
    ungroup %>%
    mutate(summarized_offense_description=fct_reorder(summarized_offense_description,n)) %>%
    ggplot() +
    geom_tile(aes(start_hour,summarized_offense_description,fill=n)) +
    scale_fill_viridis(name="# crimes") +
labs(x="Hour of the day",y="type of crime",
     title='Crime types during the day for Seattle')    

```

This shows that certain types of crimes are concentrated at certain moments of the day.

# Analysis of San Francisco

The data of San Francisco  similar, but the types of crimes are different.


```{r}
sf <- readr::read_csv("./sanfrancisco_incidents_summer_2014.csv") %>%
    clean_names()

glimpse(sf,width=120)
sf <- 
    sf %>% mutate(start_hour = hour(time))

```


```{r}
sf %>%
    count(category) %>%
    ggplot() +
    geom_col(aes(x=fct_reorder(category,n) ,y=n)) + coord_flip()
``` 


```{r}
sf %>% count(category,start_hour) %>%
    mutate(total=sum(n)) %>%
    ungroup %>%
    mutate(category=fct_reorder(category,n)) %>%
    ggplot() +
    geom_tile(aes(start_hour,category,fill=n)) +
    scale_fill_viridis(name="# crimes") +
labs(x="Hour of the day",y="type of crime",
     title="Crime types during the day for San Francisco")    
```

## Compare crime numbers for both cities
  
As the crime categories are different between both cities, I decided to compare only the over all number of crimes during the day.  



```{r}
both <- count(seattle,start_hour) %>% mutate(city="Seattle") %>% bind_rows(count(sf,start_hour) %>% mutate(city="San Francisco"))
ggplot(both) + geom_line(aes(start_hour,n,color=city)) +
    labs(x="Hour of the day",
         y="# crimes",
         title="# crimes during the day")

``` 

This shows, that there is a similar over all trend in both cities. The peak of crime is at midnight, then it decreases up to 5 o'clock, where it has the  minimum. 

The it increase again up to 12 o'clock, where it reaches an other maximum as high a the nightly maximum.

Overall Seattle has more crimes then San Francisco.

This plot shows this more clear, but the difference is small.



```{r}

bind_rows(count(seattle),count(sf),.id='city') %>% mutate(city=fct_recode(city,Seattle="1",`San Francisco`="2")) %>% ggplot() + geom_col(aes(x=city,y=n)) + labs(x="City",y="# crimes")

``` 
